---
title: "Maze binomial GLMM analyses"
output: html_document
---

This R Markdown file reproduces the three binomial generalized linear mixed-model (GLMM) analyses of maze behavior reported in the manuscript. Models were fitted with `glmmTMB`. Treatment, centered trial number, and their interaction were fitted as fixed effects; colony and bee identity were fitted as random intercepts; and repeated measurements across trials within bees were modeled with an AR(1) correlation structure.

```{r setup, message=FALSE}
library(readr)
library(glmmTMB)
```

# 1. Proportion of direction changes (Fig. 4c)

For each bee and trial, the response is the number of direction changes (`nSwitch`) out of the total number of left and right turns.

```{r direction-changes}
PropSwitch <- read_csv("../data/PropSwitch.csv", show_col_types = FALSE)

PropSwitch$Treatment <- factor(PropSwitch$Treatment)
PropSwitch$Colony    <- factor(PropSwitch$Colony)
PropSwitch$BeeID     <- factor(PropSwitch$BeeID)
PropSwitch$Trial     <- as.numeric(PropSwitch$Trial)

PropSwitch$totalTurns <- PropSwitch$nL + PropSwitch$nR

PropSwitch$TrialF <- factor(PropSwitch$Trial)
PropSwitch$Trial_c <- PropSwitch$Trial -
  mean(PropSwitch$Trial, na.rm = TRUE)

m_switch <- glmmTMB(
  cbind(nSwitch, totalTurns - nSwitch) ~
    Treatment * Trial_c +
    (1 | Colony) +
    (1 | BeeID) +
    ar1(TrialF + 0 | BeeID),
  data = PropSwitch,
  family = binomial()
)

summary(m_switch)
```

# 2. Proportion of transitions to an adjacent arm (Fig. 4d)

For each bee and trial, the response is the number of transitions to an adjacent arm (`n1Switch`) out of the total number of transitions (`nTotalSwitch`).

```{r adjacent-arm}
df <- read_csv("../data/Prop1stStep.csv", show_col_types = FALSE)

df$Treatment <- factor(df$Treatment)
df$Colony    <- factor(df$Colony)
df$BeeID     <- factor(df$BeeID)
df$Trial     <- as.numeric(df$Trial)

# Retain observations required for the model and ensure valid binomial counts.
df <- df[
  complete.cases(
    df[, c("Treatment", "Colony", "BeeID", "Trial",
           "n1Switch", "nTotalSwitch")]
  ),
]

df$n1Switch     <- round(df$n1Switch)
df$nTotalSwitch <- round(df$nTotalSwitch)

df <- subset(
  df,
  nTotalSwitch > 0 &
    n1Switch >= 0 &
    n1Switch <= nTotalSwitch
)

df$TrialF <- factor(df$Trial)
df$Trial_c <- df$Trial - mean(df$Trial, na.rm = TRUE)

m_adjacent <- glmmTMB(
  cbind(n1Switch, nTotalSwitch - n1Switch) ~
    Treatment * Trial_c +
    (1 | Colony) +
    (1 | BeeID) +
    ar1(TrialF + 0 | BeeID),
  data = df,
  family = binomial()
)

summary(m_adjacent)
```

# 3. Proportion of arm entries in which bees turned back before reaching the feeder

For each bee and trial, the response is the number of arm entries in which the bee turned back before reaching the feeder (`nRegret`) relative to the number of entries in which the feeder was reached (`nReach`). The analysis uses trials in which all eight distinct arms were completed, as in the original analysis script.

```{r turned-back}
PropRegret <- read_csv("../data/PropRegret.csv", show_col_types = FALSE)

# Rename response columns for easier coding.
names(PropRegret)[
  names(PropRegret) == "Num not reach feeder 9"
] <- "nRegret"
names(PropRegret)[
  names(PropRegret) == "Num reach feeder 9"
] <- "nReach"

# Keep only valid trials: bees completed all 8 distinct arms.
PropRegret <- subset(PropRegret, NumDistinctArms == 8)

PropRegret$Treatment <- factor(PropRegret$Treatment)
PropRegret$Colony    <- factor(PropRegret$Colony)
PropRegret$BeeID     <- factor(PropRegret$BeeID)
PropRegret$Trial     <- as.numeric(PropRegret$Trial)

PropRegret$TrialF <- factor(PropRegret$Trial)
PropRegret$Trial_c <- PropRegret$Trial -
  mean(PropRegret$Trial, na.rm = TRUE)

m_regret <- glmmTMB(
  cbind(nRegret, nReach) ~
    Treatment * Trial_c +
    (1 | Colony) +
    (1 | BeeID) +
    ar1(TrialF + 0 | BeeID),
  data = PropRegret,
  family = binomial()
)

summary(m_regret)
```

# Session information

```{r session-info}
sessionInfo()
```
